A White Noise Approach to Evolutionary Ecology

### 2020-04-18 .A White Noise Approach To Evolutionary Ecology/draft.tex
final draft

\begin{document}
\maketitle
\begin{abstract}
We derive the dynamics of the distribution of a quantitative character
and the abundance of a biological population from a stochastic partial
differential equation driven by space-time white noise. In the process
we develop a useful set of heuristics to operationalize the powerful,
but abstract theory of white noise and measure-valued stochastic
processes. This approach allows us to compute the full implications of
demographic stochasticity on phenotypic distributions and abundances of
populations. We demonstrate the utility of our approach by deriving a
quantitative genetic model of diffuse coevolution mediated by
exploitative competition for a continuum of resources. Other than trait
and abundance distributions, this model predicts interaction networks
parameterized by rates of interactions, competition coefficients, and
selection gradients. We briefly investigate the relationship between
selection gradients and competition coefficients. This illustrative
investigation suggests selection gradients can be either positively or
negatively correlated with competition coefficients depending on the
ratio of interspecific trait variation to intraspecific trait variation.
Hence, this approach contributes to the development of a synthetic
theory of evolutionary ecology by formalizing first principle
derivations of stochastic models that underlie rigorous investigations
of the relationship between feedbacks of biological processes and the
patterns of diversity they produce.
\end{abstract}

@@

Current theoretical approaches to synthesize evolution and ecology have
capitalized on the fact that biological fitness plays a key role in
determining both sets of dynamics. While correlation of fitness and
genotype is the basis of evolution by natural selection, the mean
fitness across all individuals in a population determines the growth,
stasis or decline of abundance. In section \ref{deterministic} we review
the mathematical formalization of this connection, which has been
established in the contexts of population genetics (Crow and Kimura
1970; Roughgarden 1979), evolutionary game theory (Hofbauer and Sigmund
1998; Nowak 2006; Lion 2018), quantitative genetics (Lande 1982; Doebeli
1996; Lion 2018) and a unifying framework for these three distinct
approaches to evolutionary theory (Champagnat, Ferrière, and Méléard
2006) which includes our approach as a special case. We note here there
are two primary differences between our aims and those of Champagnat,
Ferrière, and Méléard (2006). First, instead of developing a unifying
approach to evolutionary theory, we focus on developing a stochastic
synthesis of population dynamics and quantitative genetics including
specific expressions for the dynamics of abundance, mean trait and
additive genetic variance. Second, although Champagnat, Ferrière, and
Méléard (2006) provide rigorous proofs of results relevant to our
approach, they do not communicate the nessecary tools we make use of for
deriving the dynamics of abundance, mean trait and additive genetic
variance. We aim to communicate these tools to a wide audience by
providing an intuitive, yet accurate treatment of white noise and
associated heuristics useful for deriving new stochastic models of
evolutionary ecology.

Reviewing the above accomplishments reveals a beautiful synthesis of
evolution and population ecology. However, it also reveals a gap in
theoretical approaches to incorporate the intrinsically random nature of
populations. Specifically, in theoretical quantitative genetics the
derivation of a population's response to random genetic drift is derived
in discrete time under the assumption of constant effective population
size using arguments based on properties of random samples (Lande 1976).
Though this approach conveniently mimics the formalism provided by the
Wright-Fisher model of population genetics, real population sizes
fluctuate over time. Furthermore, since these fluctuations are
themselves stochastic, it seems natural to derive expressions for the
evolutionary response to demographic stochasticity and consider how the
results relate to characterizations of random genetic drift. This can be
done in continuous time for population genetic models without too much
technical overhead, assuming a finite number of alleles (Lande, Engen,
and Sæther 2009; Parsons, Quince, and Plotkin 2010; Gomulkiewicz, Krone,
and Remien 2017). However, for populations with a continuum of types,
such as a quantitative trait, finding a formal approach to derive the
evolutionary response to demographic stochasticity has remained a vexing
mathematical challenge. Previous work integrating stochastic processes
with quantitative genetic models have profited from the simplifying
assumption that phenotypic and genetic variances remain constant over
time, yeilding analytical approximations for the joint dynamics of
abundance and the mean of a quantitative trait (Engen, Lande, and Sæther
2013). These results not only ignore the evolution genetic variance, but
they focus on environmental stochasticity and assume sufficiently large
populations so that demographic stochasticity is negligible. Here we
account for both demographic stochasticity and the evolution of additive
genetic variance by combining the calculus of white noise with results
on rescaled limits of branching Brownian motion processes (BBM) and
stochastic partial differential equations (SPDE), which are stochastic
analogs of deterministic partial differential equations (PDE). Our goal
has two components: 1) Establish a novel synthetic approach to
theoretical evolutionary ecology that provides a formal connection
between demographic stochasticity and random genetic drift in the
context of quantitative traits. 2) Communicate some useful properties of
space-time white noise, BBM and SPDE to as wide of audience as possible.
With this goal in mind we will not provide a rigorous treatment of any
of these deep subjects. Instead, we introduce a set of heuristics that
only require the basic concepts of Reimann integration, partial
differentiation and some exposure to Brownian motion and stochastic
ordinary differential equations (SDE). For a concise introduction to SDE
and Brownian motion, we recommend the primer by Evans (2014). Rigorous
treatments of SPDE and rescaled limits of BBM can be found in Walsh
(1986) and Dawson (1993) respectively.

Through accomplishing these goals this work yields five main results.
Firstly, we develop a concise, but accurate approach for understanding
SPDE and provide heuristics to empower mathematical evolutionary
ecologists to derive novel models that relax key assumptions commonly
made in combining models of evolutionary and abundance dynamics. Second,
we formally derive general SDE that express the joint dynamics of
abundance, mean trait and additive genetic variance for populations
exhibiting a wide range of potential phenotypic distributions. Third, we
simplify these SDE by approximating the abundance density of phenotypes
with a Gaussian curve and integrate the results with a model of
imperfect inheritance to find formula that generalize classical
quantitative genetic expressions. This includes an expression for
additive genetic variance analogous to the expression for the response
in mean phenotype found by Lande (1976). Fourth, following classical
niche theory we provide a rigorous mechanistic derivation of fitness for
populations experiencing resource competition and abiotic stabilizing
selection using diffusion limits of measure-valued branching processes.
This calculation provides a demonstration for performing such diffusion
limits that can be applied in other contexts. Fifth, we explore
implications of the resulting model of diffuse coevolution by
investigating the relationship between selection gradients and
competition coefficients.

We now summarize the basic structure of the paper. To provide motivation
for the stochastic equations developed later and background for our
model of coevolution, we begin with \S\ref{deterministic} by briefly
summarizing derivations of deterministic dynamics of biological
populations. Starting with a PDE (which we refer to as the Deterministic
Asexual Gaussian allelic model with Abundance dynamics, abbreviated
DAGA), we arrive at a general set of ordinary differential equations
modelling the dynamics of abundance, trait mean and trait variance. We
discuss the relationship between DAGA and previous models of phenotypic
evolution that have been employed in theoretical quantitative genetics.
Noting that a natural stochastic generalization of DAGA has been
previously derived, we find a path to derive SDE describing the
evolutionary response to demographic stochasticity. We accomplish this
in \S\ref{wnc} by discussing how a diffusion limit of a spatially
structured branching process (i.e., a measure-valued branching process)
leads to the natural SPDE appropriate for our study. We refer to this
SPDE as the Stochastic Asexual Gaussian allelic model with Abundance
dynamics (abbreviated SAGA). The diffusion limit in turn provides a
rigorous method for constructing fitness functions used in models of
evolutionary ecology. However, before we discuss measure-valued braching
processes and their diffusion limits, we find it necessary to review
some fundamental results on the calculus of white noise and introducing
a set of mathematical tools based on these results. In \S\ref{wnc} and
SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE
generalizing our deterministic results to account for demographic
stochasticity. We find that the effect of demographic stochasticity on
the evolution of mean trait and phenotypic variance characterizes the
process of random genetic drift in continuous time. However, although
biologically insightful, these equations remain difficult to analyze and
implement numerically. In \S\ref{equations} we use an assumption of
normally distributed trait values to simplify these expressions into
formulae that are much easier to work with. We then account for the
constraint of adaptive evolution on the amount of heritable variation in
a population by extending these results via a model of imperfect
inheritance. The resulting equations coincide with classical results in
quantitative genetics as a special case. In \S\ref{coev} we combine the
derived equations of population dynamics with classical niche theory to
formulate a model of coevolution across a guild of \(S\) species
participating in exploitative competition along a common resource
continuum. In SM \S\ref{diffuse} we apply a classical theorem on
rescaled limits of BBM that allow for ecological interactions to provide
a rigorous derivation. To gain biological insight, in \S\ref{dynamics}
we numerically integrate our model of coevolution for \(S=100\) species,
tracking the dynamics of traits and abundances under scenarios of weak
and strong competition. We include an account of the natural history of
the simulated system and discuss the significance of demographic
stochasticity for structuring ecological communities. In \S\ref{ecoevo}
we provide expressions for selection gradients and competition
coefficients implied by our model and use these expressions to
investigate the relationship between the degree of competition and
strength of coevolution. Finally, \S\ref{conclusion} concludes with a
summary of accomplishments, a few remarks on the limits of this approach
and future directions to incorporate more explicitly the effects of
small populations, the genetic architecture of quantitative traits,
feedbacks with ecosystem processes and the macroevolutionary history of
interacting lineages.

@@

At the core of our approach is a stochastic analog of the replicator
equation with mutation in continuous time and phenotypic space (Taylor
and Jonker 1978; Schuster and Sigmund 1983). From this stochastic
replicator-mutator equation we derive a system of SDE for the dynamics
of abundance, mean trait and additive genetic variance of a population.
Hence, our approach develops a quantitative genetic theory of
evolutionary ecology. A popular alternative to quantitative genetics is
the theory of adaptive dynamics. As demonstrated by Page and Nowak
(2002) and Champagnat, Ferrière, and Méléard (2006), the canonical
equation of adaptive dynamics can be derived from the replicator-mutator
equation. Thus, one could start from the atomic roots of our approach
and pursue a stochastic adaptive dynamic theory instead. We choose the
former in anticipation of an extension of our approach that explicitly
models the genetic details of populations.

In this section we review the derivations of the replicator-mutator
equation and trait dynamics from abundance dynamics and extend these
formulae along with related results to the stochastic case. The results
established in this section provide the framework from which larger
scale ecological stuctures, such as species abundance distributions and
interaction networks, can be computed.

@@

Our review begins by considering the dynamics of an asexually
reproducing population in a homogeneous environment. For simplicity, we
first assume individuals are haploid and carry one of \(K\) alleles each
with a different fitness expressed as growth rate before introducing a
model involving a quantitative trait. Under these assumptions, the
derivation of the evolution of allele frequencies due to natural
selection can be derived from expressions of exponential growth. This,
and a few related approaches, have been provided by Crow and Kimura
(1970, \S 5.3). Specifically, denoting \(\nu_i\) the abundance of
individuals with allele \(i\) and \(m_i\) the growth rate of allele
\(i\) (called the Malthusian parameter in Crow and Kimura 1970), we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation} Starting from this model, we get the total abundance of
the population as \(N=\sum_{i=1}^K\nu_i\), the frequency of allele \(i\)
as \(p_i=\nu_i/N\) and the mean fitness of the population as
\(\bar m=\sum_{i=1}^Kp_im_i\). Hence, we can employ some elementary
calculus to derive the dynamics of abundance \(dN/dt\) and the dynamics
of allele frequencies \(dp_1/dt,\dots,dp_K/dt\) as \begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation} \begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation} Two important observations of these equations include 1)
mean fitness \(\bar m\) determines the abundance dynamics of the entire
population and 2) allele \(i\) will increase (decrease) in frequency if
\(m_i>\bar m\) (\(<\bar m\)). Equation (\ref{popgen}) is known in the
field of evolutionary game theory as the replicator equation (Hofbauer
and Sigmund 1998; Nowak 2006; Lion 2018; Taylor and Jonker 1978;
Schuster and Sigmund 1983). Instead of being explicitly focused on
alleles, the replicator equation describes the fluctuations of relative
abundances of various \emph{types} in a population in terms of the vital
rates of each type. Using a matrix of transition rates between differing
types, it is straight-forward to extend the replicator equation to
include mutation, which is known as the replicator-mutator equation
(Nowak 2006).

@@

To stay within the realm of biological plausibility we require a set
technical assumptions. First, we assume the initial abundance density is
continuous, non-negative, integrable and has finite trait mean and
variance. That is, we assume \(\nu(x,0)\) is continuous in \(x\),
satisfies \(\nu(x,0)\geq0\) for all \(x\in\mathbb{R}\) and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation} \begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation} \begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation} where \(\bar x(t)\) and \(\sigma^2(t)\) are respectively
the mean trait and phenotypic variance at time \(t\geq0\). Second, we
assume selection is determined by the growth rate \(m(h,x)\) that is
differentiable with respect to both arguments and is bounded above by
some \(r\in\mathbb{R}\) for all \(x\in\mathbb{R}\) and \(h\geq0\). We
denote by \(K\) an operator that takes functions such as \(\nu(x,t)\) as
its argument and returns other functions that may still depend on the
spatial variable \(x\) and the temporal variable \(t\). For our
application we will be focused on operators \(K\) that can be written as
\((K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy\) for some
non-negative and bounded function \(\kappa\). We set \(h=(K\nu)(x,t)\)
to account for nonlocal effects of the abundance density \(\nu(x,t)\) on
the fitness of individuals with trait value \(x\) (Volpert 2014). As an
example of a nonlocal effect, consider a population experiencing
intraspecific competition for a common resource such that the
competitive interactions are not mediated by any set of traits. In this
case all individuals in the population effect the fitness of one another
even though they may exhibit a diverse range of trait values. The term
nonlocal comes from interpreting the trait value as spatial location.
From here on we abbreviate \(m(h,x)=m\big((K\nu)(x,t),x\big)\) to just
\(m(\nu,x)\). Third, we assume mutation is captured by diffusion with
coefficient \(\frac{\mu}{2}\). With these technicalities addressed we
model the demographic dynamics of a population and the dynamics of a
quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation} with the initial condition \(\nu(x,0)\) described above.
This PDE is semilinear due to the dependency of the growth rate
\(m(\nu,x)\) on the solution \(\nu(x,t)\) and is referred to as a scalar
reaction-diffusion equation (Zheng 2004; Evans 2010). To ensure
existence and uniqueness of solutions to (\ref{eq1}), we further assume
that for each \(M>0\) there exists a constant \(L_M>0\) depending on
\(M\) such that
\(\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M\) implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}\nu_1(x)-\nu_2(x)dx.
\end{equation} In SM \S\ref{finite} we combine the above assumptions on
growth rate, mutation and initial condition to prove that solutions to
(\ref{eq1}) satisfy \(N(t),|\bar x(t)|,\sigma^2(t)<+\infty\) for all
\(t\geq0\).

>>>>

\begin{document}
\maketitle
\begin{abstract}
We derive the dynamics of the distribution of a quantitative character
and the abundance of a biological population from a stochastic partial
differential equation driven by space-time white noise. In the process
we develop a useful set of heuristics to operationalize the powerful,
but abstract theory of white noise and measure-valued stochastic
processes. This approach allows us to compute the full implications of
demographic stochasticity on phenotypic distributions and abundances of
populations. We demonstrate the utility of our approach by deriving a
quantitative genetic model of diffuse coevolution mediated by
exploitative competition for a continuum of resources. Other than trait
and abundance distributions, this model predicts interaction networks
parameterized by rates of interactions, competition coefficients, and
selection gradients. We briefly investigate the relationship between
selection gradients and competition coefficients. This illustrative
investigation suggests absolute values of linear selection gradients are
negatively correlated with competition coefficients when the ratio of
interspecific trait variation to intraspecific trait variation is
approximately less than \(1.25\) and positively correlated otherwise.
The inverse relationship holds for quadratic selection gradients except
with threshold diversity ratio equal to unity instead of \(1.25\).
Hence, this approach contributes to the development of a synthetic
theory of evolutionary ecology by formalizing first principle
derivations of stochastic models that underlie rigorous investigations
of the relationship between feedbacks of biological processes and the
patterns of diversity they produce.
\end{abstract}

@@

Current theoretical approaches to synthesize evolution and ecology have
capitalized on the fact that biological fitness plays a key role in
determining both sets of dynamics. While correlation of fitness and
genotype is the basis of evolution by natural selection, the mean
fitness across all individuals in a population determines the growth,
stasis or decline of abundance. In section \ref{deterministic} we review
the mathematical formalization of this connection, which has been
established in the contexts of population genetics (Crow and Kimura
1970; Roughgarden 1979), evolutionary game theory (Hofbauer and Sigmund
1998; Nowak 2006; Lion 2018), quantitative genetics (Lande 1982; Doebeli
1996; Lion 2018) and a unifying framework for these three distinct
approaches to evolutionary theory (Champagnat, Ferrière, and Méléard
2006) which includes our approach as a special case. We note here there
are two primary differences between the results presented in this paper
and those of Champagnat, Ferrière and Méléard (2006). First, instead of
developing a unifying approach to evolutionary theory, we focus on
developing a stochastic synthesis of population dynamics and
quantitative genetics including specific expressions for the dynamics of
abundance, mean trait and additive genetic variance. Second, although
Champagnat, Ferrière and Méléard (2006) provide rigorous proofs of
results relevant to our approach, they do not communicate the nessecary
tools we make use of for deriving the dynamics of abundance, mean trait
and additive genetic variance. We aim to communicate these tools to a
wide audience by striking a balance between intuitive appeal and
mathematical rigor.

Although reviewing previously established theoretical results reveals a
beautiful synthesis of evolution and population ecology, it also reveals
a gap in theoretical approaches to incorporate the intrinsically random
nature of populations. Specifically, in theoretical quantitative
genetics the derivation of a population's response to random genetic
drift is derived in discrete time under the assumption of constant
effective population size using arguments based on properties of random
samples (Lande 1976). Though this approach conveniently mimics the
formalism provided by the Wright-Fisher model of population genetics,
real population sizes fluctuate over time. Furthermore, since these
fluctuations are themselves stochastic, it seems natural to derive
expressions for the evolutionary response to demographic stochasticity
and consider how the results relate to characterizations of random
genetic drift. This can be done in continuous time for population
genetic models without too much technical overhead, assuming a finite
number of alleles (Lande, Engen, and Sæther 2009; Parsons, Quince, and
Plotkin 2010; Gomulkiewicz, Krone, and Remien 2017). However, for
populations with a continuum of types, such as a quantitative trait,
finding a formal approach to derive the evolutionary response to
demographic stochasticity has remained a vexing mathematical challenge.
In this paper we close this gap by combining the calculus of white noise
with results on rescaled limits of branching Brownian motion processes
(BBM) and stochastic partial differential equations (SPDE), which are
stochastic analogs of deterministic partial differential equations
(PDE). Our goal has two components: 1) Establish a novel synthetic
approach to theoretical evolutionary ecology that provides a formal
connection between demographic stochasticity and random genetic drift in
the context of quantitative traits. 2) Communicate some useful
properties of space-time white noise, BBM and SPDE to as wide of
audience as possible. With this goal in mind we will not provide a
rigorous treatment of any of these deep subjects. Instead, we introduce
a set of heuristics that only require the basic concepts of Reimann
integration, partial differentiation and some exposure to Brownian
motion and stochastic ordinary differential equations (SDE). For a
concise introduction to SDE and Brownian motion, we recommend the primer
by Evans (2014). Rigorous treatments of SPDE and rescaled limits of BBM
can be found in Walsh (1986) and Dawson (1993) respectively.

Through accomplishing these goals this work yields five main results.
Firstly, we develop a concise, but accurate approach for understanding
SPDE and provide heuristics to empower mathematical evolutionary
ecologists to derive novel models that relax key assumptions commonly
made in combining models of evolutionary and abundance dynamics. Second,
we formally derive general SDE driven by demographic stochasticity that
express the joint dynamics of abundance, mean trait and additive genetic
variance for populations exhibiting a wide range of potential phenotypic
distributions. Although these dynamics have been considered before while
accounting for demographic stochasticity (Polechová and Barton 2015), to
our knowledge this is the first time they have been rigorously derived
from first principles. Third, we simplify these SDE by approximating the
abundance density of phenotypes with a Gaussian curve and integrate the
results with a model of imperfect inheritance to find formula that
generalize classical quantitative genetic expressions. This includes an
expression for the response of additive genetic variance to selection in
terms of fitness gradients analogous to the expression for the response
in mean phenotype found by Lande (1976). Fourth, following classical
niche theory we provide a rigorous mechanistic derivation of fitness for
populations experiencing resource competition and abiotic stabilizing
selection using diffusion limits of measure-valued branching processes.
This calculation provides a demonstration for performing such diffusion
limits that can be applied in other contexts. Fifth, we explore
implications of the resulting model of diffuse coevolution by
investigating the relationship between selection gradients and
competition coefficients, discovering the conditions under which these
quantities positively or negatively covary.

We now summarize the basic structure of the paper. To provide motivation
for the stochastic equations developed later and background for our
model of coevolution, we begin with \S\ref{deterministic} by briefly
summarizing derivations of deterministic evolutionary and demographic
dynamics. Starting with a PDE (which we refer to as the Deterministic
Asexual Gaussian allelic model with Abundance dynamics, abbreviated
DAGA), we arrive at a general set of ordinary differential equations
modelling the dynamics of abundance, trait mean and trait variance. We
discuss the relationship between DAGA and previous models of phenotypic
evolution that have been employed in theoretical quantitative genetics.
In \S\ref{wnc} we discuss how a diffusion limit of a spatially
structured branching process (i.e., a measure-valued branching process)
leads to a SPDE that provides a natural stochastic generalization of
DAGA. We refer to this SPDE as the Stochastic Asexual Gaussian allelic
model with Abundance dynamics (abbreviated SAGA). The diffusion limit in
turn provides a rigorous method for constructing fitness functions used
in models of evolutionary ecology. However, before we discuss
measure-valued braching processes and their diffusion limits, we find it
necessary to review some fundamental results on the calculus of white
noise and introduce a set of mathematical tools based on these results.
In SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE
generalizing our deterministic results to account for demographic
stochasticity. We find the effect of demographic stochasticity on the
evolution of mean trait and phenotypic variance characterizes the
process of random genetic drift in continuous time. However, although
biologically insightful, these equations remain difficult to analyze and
implement numerically. In \S\ref{equations} we use an assumption of
normally distributed trait values to simplify these expressions into
tractable formula that express the evolutionary response to selection in
terms of fitness gradients. We then account for the constraint of
adaptive evolution on the amount of heritable variation available in a
population by extending these results via a model of imperfect
inheritance. The resulting equations coincide with classical results in
quantitative genetics as a special case. In \S\ref{coev} we combine the
derived equations of population dynamics with classical niche theory to
formulate a model of coevolution across a guild of \(S\) species
participating in exploitative competition along a common resource
continuum. In SM \S\ref{diffuse} we apply a result on rescaled limits of
BBM to provide a rigorous derivation. In \S\ref{dynamics} we illustrate
our model of coevolution through numerical integration for \(S=100\)
species, tracking the dynamics of traits and abundances under scenarios
of weak and strong competition. We include an account of the natural
history of the simulated system and discuss the significance of
demographic stochasticity for structuring ecological communities. In
\S\ref{ecoevo} we provide expressions for selection gradients and
competition coefficients implied by our model and use these expressions
to investigate the relationship between the degree of competition and
strength of coevolution. Finally, \S\ref{conclusion} concludes with a
summary of accomplishments, a few remarks on the limits of this approach
and future directions to incorporate more explicitly the effects of
small populations, the genetic architecture of quantitative traits,
feedbacks with ecosystem processes and the macroevolutionary history of
interacting lineages.

@@

At the core of our approach is a stochastic analog of the replicator
equation with mutation in continuous time and phenotypic space (Taylor
and Jonker 1978; Schuster and Sigmund 1983). From this stochastic
replicator-mutator equation we derive a system of SDE for the dynamics
of abundance, mean trait and additive genetic variance of a population.
Hence, our approach develops a quantitative genetic theory of
evolutionary ecology. A popular alternative to quantitative genetics is
the theory of adaptive dynamics. As demonstrated by Page and Nowak
(2002) and Champagnat, Ferrière and Méléard (2006), the canonical
equation of adaptive dynamics can be derived from the replicator-mutator
equation. Thus, one could start from the atomic roots of our approach
and pursue a stochastic adaptive dynamic theory instead. We choose the
former in anticipation of an extension of our approach that explicitly
models the genetic details of populations.

In this section we review the derivations of the replicator-mutator
equation and trait dynamics from abundance dynamics and extend these
formula along with related results to the stochastic case. The results
established in this section provide the framework from which larger
scale ecological stuctures, such as species abundance distributions and
interaction networks, can be computed.

@@

Our review begins by considering the dynamics of an asexually
reproducing population in a homogeneous environment. For simplicity, we
first assume individuals are haploid and carry one of \(K\) alleles each
with a different fitness expressed as growth rate before introducing a
model involving a quantitative trait. Under these assumptions, the
derivation of the evolution of allele frequencies due to natural
selection can be derived from expressions of exponential growth. This,
and a few related approaches, have been provided by Crow and Kimura
(1970, \S 5.3). Specifically, denoting \(\nu_i\) the abundance of
individuals with allele \(i\) and \(m_i\) the growth rate of allele
\(i\) (called the Malthusian parameter in Crow and Kimura 1970), we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation} Starting from this model, we get the total abundance of
the population as \(N=\sum_{i=1}^K\nu_i\), the frequency of allele \(i\)
as \(p_i=\nu_i/N\) and the mean fitness of the population as
\(\bar m=\sum_{i=1}^Kp_im_i\). Hence, we can employ some elementary
calculus to derive the dynamics of abundance \(dN/dt\) and the dynamics
of allele frequencies \(dp_1/dt,\dots,dp_K/dt\) as \begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation} \begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation} Two important observations of these equations include (i)
mean fitness \(\bar m\) determines the abundance dynamics of the entire
population and (ii) allele \(i\) will increase (decrease) in frequency
if \(m_i>\bar m\) (\(<\bar m\)). Equation (\ref{popgen}) is known in the
field of evolutionary game theory as the replicator equation (Hofbauer
and Sigmund 1998; Nowak 2006; Lion 2018; Taylor and Jonker 1978;
Schuster and Sigmund 1983). Instead of being explicitly focused on
alleles, the replicator equation describes the fluctuations of relative
abundances of various \emph{types} in a population in terms of the vital
rates of each type. Using a matrix of transition rates between differing
types, it is straight-forward to extend the replicator equation to
include mutation, which is known as the replicator-mutator equation
(Nowak 2006).

@@

To stay within the realm of biological plausibility we require a set
technical assumptions. First, we assume the initial abundance density is
continuous, non-negative, integrable and has finite trait mean and
variance. That is, we assume \(\nu(x,0)\) is continuous in \(x\),
satisfies \(\nu(x,0)\geq0\) for all \(x\in\mathbb{R}\) and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation} \begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation} \begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation} where \(\bar x(t)\) and \(\sigma^2(t)\) are respectively
the mean trait and phenotypic variance at time \(t\geq0\). Second, we
assume selection is determined by the growth rate \(m(h,x)\) that is
differentiable with respect to both arguments and is bounded above by
some \(r\in\mathbb{R}\) for all \(x\in\mathbb{R}\) and \(h\geq0\). We
also assume \((1+h)C\leq m(h,x)\) for some constant \(C>0\) so that
\(m(h,x)\) is linear or more than linear in \(h\). We denote by \(K\) an
operator that takes functions such as \(\nu(x,t)\) as its argument and
returns other functions that may still depend on the spatial variable
\(x\) and the temporal variable \(t\). For our application we will be
focused on operators \(K\) that can be written as
\((K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy\) for some
non-negative and bounded function \(\kappa\). We set \(h=(K\nu)(x,t)\)
to account for nonlocal effects of the abundance density \(\nu(x,t)\) on
the fitness of individuals with trait value \(x\) (Volpert 2014). As an
example of a nonlocal effect, consider a population experiencing
intraspecific competition for a common resource such that the
competitive interactions are not mediated by any set of traits. In this
case all individuals in the population effect the fitness of one another
even though they may exhibit a diverse range of trait values. The term
nonlocal comes from interpreting the trait value as spatial location.
From here on we abbreviate \(m(h,x)=m\big((K\nu)(x,t),x\big)\) to just
\(m(\nu,x)\). Third, we assume mutation is captured by diffusion with
coefficient \(\frac{\mu}{2}\). With these technicalities addressed we
model the demographic dynamics of a population and the dynamics of a
quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation} with the initial condition \(\nu(x,0)\) described above.
Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as
a scalar reaction-diffusion equation (Zheng 2004; Evans 2010). To ensure
existence and uniqueness of solutions to (\ref{eq1}), we further assume
that for each \(M>0\) there exists a constant \(L_M>0\) depending on
\(M\) such that
\(\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M\) implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}|\nu_1(x)-\nu_2(x)|dx.
\end{equation} In SM \S\ref{finite} we combine the above assumptions on
growth rate, mutation and initial condition to prove that solutions to
(\ref{eq1}) satisfy \(N(t),|\bar x(t)|,\sigma^2(t)<+\infty\) for all
\(t\geq0\).

Interpreting the trait value \(x\) as spatial location and ignoring
nonlocal effects, which can be satisfied by setting \(K\nu=\nu\), this
model of spatially distributed population dynamics has been intensely
studied (for a review, see Cantrell and Cosner 2004). When \(\mu=0\),
equation (\ref{eq1}) can be seen as an analog of equation (\ref{first})
for a continuum of types. By assuming mutation acts as diffusion, the
effect of mutation causes \(\nu(x,t)\) to flatten out over time. In
fact, if the growth rate is constant across \(x\), then this model of
mutation will cause \(\nu(x,t)\) to converge to a flat line as
\(t\to\infty\).

Although clearly an idealized representation of biological reality, this
model is sufficiently general to capture a large class of dynamics
including density dependent growth and frequency dependent selection. As
an example, logistic growth combined with quadratic stabilizing
selection can be captured using the growth rate
\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation} where \(r\in\mathbb{R}\) is the intrinsic rate of growth
in the absence of abiotic stabilizing selection, \(a>0\) is the strength
of abiotic stabilizing selection around the phenotypic optimum
\(\theta\in\mathbb{R}\), \(c>0\) captures the sensitivity of fitness to
intraspecific competition and \(\kappa(x-y)=1\) so that competitive
interactions cause the same reduction in fitness regardless of trait
value. This exemplary fitness function has a few convenient properties.
First, the effect of competition induces a local carrying capacity on
the population, leading to a finite equilibrium abundance over bounded
subsets of \(\mathbb{R}\). Second, abiotic selection prevents the
abundance density from diffusing too far from the abiotic optimum,
leading to a finite equilibrial abundance across all of \(\mathbb{R}\),
\(\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})\),
finite equilibrial mean trait \(\lim_{t\to\infty}\bar x(t)=\theta\) and
finite equilibrial phenotypic variance
\(\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}\), so long as
\(\bar x(0)\in\mathbb{R}\), \(\sigma^2(0)\in[0,+\infty)\),
\(N(0)\in(0,+\infty)\) and \(r>\frac{1}{2}\sqrt{a\mu}>0\). We prove
these statements in SM \S\ref{equilib}. Note in particular the
coincidence of equilibrial phenotypic variance predicted by this model
and that predicted by the classic Gaussian allelic model of quantitative
genetics (Lande 1975; Bürger 2000; Johnson and Barton 2005). In
\S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is
closely related to a fitness function derived from niche theory that we
use to model coevolution driven by resource competition.
